### 2020-11-03 manuscript/manuscript.rmd
fix Fig.2 and update comments from Nicole

![](figures/Figure_overview.png)
**Fig. 1.** Two alternative phylogenetic hypotheses for the root of the animal tree. (A) The Ctenophora-sister hypothesis posits that there is a clade (designated by the orange node) that includes all animals except Ctenophora, and that Ctenophora is sister to this clade. (B) The Porifera-sister hypothesis posits that there is a clade (designated by the green node) that includes all animals except Porifera, and that Porifera is sister to this clade. Testing these hypotheses requires evaluating the support for each of these alternative nodes. (C) The animals and their outgroup choice, showing the three progressively more inclusive clades Choanimalia, Holozoa, and Opisthokonta. (D) A ctenophore, commonly known as comb jellies. (E) A poriferan, commonly known as a sponge.

@@

Sampling of outgroup taxa (non-animals, in this case) is critical to phylogenetic rooting questions, since the node where the outgroup subtree attaches to the ingroup subtree is the root of the ingroup. There has therefore been extensive focus on improving outgroup sampling when testing phylogenetic hypotheses about rooting [@graham2002rooting]. Most studies addressing the animal root have removed more distantly related outgroup taxa in some analyses to explore the effect of outgroup selection to ingroup topology [@ryan2013genome; @pisani2015genomic]. Three progressively more inclusive clades have often been investigated: Choanimalia (animals plus most closely related Choanoflagellatea), Holozoa (Choanimalia plus more distantly related Holozoa), and Opisthokonta (Holozoa + Fungi).

@@

OOur summary of previous phylogenetic analyses (Fig. 3A) shows that Ctenophora-sister is supported in analyses that span the full range of outgroup sampling and models used to date, including some analyses with restricted outgroup sampling and models that accommodate site-heterogeneous equilibrium frequencies with CAT. This is consistent with previous assessments of the problem [@whelan2016let], but is drawn from a much more extensive and systematic examination. The only analyses that support Porifera-sister have reduced outgroup sampling (Choanimalia, Holozoa) and site-heterogeneous models with CAT. Model adequacy assessments generally favor GTR+CAT over Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], but because GTR+CAT is so parameter rich, many analyses that use a model with GTR+CAT do not converge. The fact that Porifera-sister is recovered only for particular models with particular outgroup sampling indicates that model and outgroup interact, and that this interaction is fundamental to understanding the range of results obtained across analyses.

@@

For each of the published studies, we selected the matrix that was the primary focus of the manuscript, or has been reanalyzed extensively in other studies, for further analysis. For each of these matrices, we progressively trimmed taxon sampling to create Opisthokonta, Holozoa, and Choanimalia versions, where permitted by original outgroup sampling. This produced `r analyses_new$matrix%>% unique()%>% length()` data matrices from 11 studies that presented new sequence data and for which partition data were available.

@@

We examined the specific impact of this large difference in category number on animal root position. It is currently not possible to use more than 60 categories for C models in IQ-TREE, but the number of categories can be set *a priori* in Phylobayes using nCAT. We therefore varied the number of categories in Phylobayes analyses (Fig. 4; Supplementary Table 5). We found that Poisson+nCAT60 analyses in PhyloBayes, like Poisson+C60 and WAG+C60 IQ-TREE analyses, provide strong support of Ctenophora-sister. This indicates that the difference in results between unconstrained CAT analyses in PhyloBayes and C60 analyses in IQ-TREE is not due to differences in the software or other implementation factors, but due to the large difference, in excess of ten-fold, in the number of site categories. When we increased the number of categories in PhyloBayes nCAT analyses, we observe the transition from support for Ctenophora-sister to an unresolved root to Porifera-sister (Fig. 4). For example, for the Whelan2017_strict matrix this transition occurs between 60 -- 120 categories when using Poisson+nCAT model. Due to computational limitations of GTR models, we only ran GTR+CAT and GTR+nCAT60 models on representative matrices with Choanimalia sampling. We found that for Whelan2017 matrices, support shifted from Porifera-sister with Poisson+CAT model to Ctenophora-sister using GTR+CAT model. Moreover, we also found all results strongly supported Ctenophora-sister with GTR+nCAT60 models.

These results further clarify when analyses support Porifera-sister (*e.g.* Whelan2017: Extended Data Fig. 1; Philippe2009: Extended Data Fig. 2): when outgroup sampling is restricted (Choanimalia), when a Poisson (rather than a GTR) exchange matrix is used, and when a very large number of site categories is used (unconstrained CAT, giving hundreds of equilibrium frequency categories). Analyses under other conditions either support Ctenophora-sister or are unresolved. This is consistent across published analyses and our new panels of analyses. The question, then, isn’t why similar analyses give different results, but how we should interpret variation in results when we run analyses that differ in these specific respects. If we fix the first two features to conditions that are necessary for Porifera-sister support (Choanimalia taxon sampling and a Poisson exchange matrix), there are several insights that we can glean from examining how the number of equilibrium frequency categories impacts results that sheds light on interpretation of those results.

The first questions regard the relative fit of adding aditional site-heterogeneous categories, and the adequacy of models with different numbers of categories. ModelFinder selects site heterogeneous C60 models (Supplementary Table 3), but IQ-TREE often gives a warning under C60 that the model may overfit, with too many categories, because some mixture weights are close to 0 (Supplementary Table 6). While ModelFinder considers the relative fit of models, this is distinct from model adequacy -- an absolute measure of how well the model can describe the variation in the data. Cross-Validation (CV) is an important tool to assess adequacy by examining the predictive power of the model, *i.e.* the ability of the model to predict observations [@lartillot2009phylobayes]. It could be that a model with a very large number of categories are required to adequately describe the site heterogeneity in these data, and that we should therefore favor the Porifera-sister results obtained in unconstrained CAT analyses. To evaluate the predictive performance and model-fit between Poisson+CAT and Poisson+nCAT60 models in Whelan2017_strict and Philippe2009_Choanimalia matrices, we conducted Bayesian cross-validation experiments with PhyloBayes. We found nearly identical distributions of cross-validation scores for Poisson+nCAT60 and Poisson+CAT models (Extended Data Fig. 5). A t-test analysis shows that there is no significant difference between them for either matrix (Whelan2017_strict: $p-value=`r P_value_Whelan2017_strict`$; Philippe2009_Choanimalia: $p-value=`r P_value_Philippe2009`$). This indicates that 60 categories in a Poisson+nCAT60 model are as adequate as at predicting the data as the hundreds of categories in an unconstrained Poisson+CAT model. This suggests there is no reason to favor these narrow analysis conditions under which Porifera-sister are obtained.

In a site-heterogeneous model, each site is allocated to one of the categories in each sample. Different categories can have very different numbers of sites. We examine category allocations in the last chain samples of analyses of the Philippe2009_Choanimalia and Whelan2017_strict matrices (Extended Data Fig. 6A-B), which have `r n_categories_phil_cat` and `r n_categories_whel_cat` categories, respectively. The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009_Choanimalia matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix. This long tail of categories that apply to a very small fraction of sites is in stark contrast to nCAT60 analyses, constrained to 60 categories, which have no such long tail of rare categories (Extended Data Fig. 6A-B). The fact that so many categories apply to such a small fraction of data be help explain why increasing the number of categories more than ten-fold has so little impact on the predictive power of these far more complex models.

@@

Although a higher proportion of genes support Ctenophora-sister with site-heterogeneous C60 models, the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models. In an extreme case, in matrices from Ryan2013_est nearly 30% of genes changed from strong $\Delta$|lnL|>2) to weak Ctenophora-sister signal ($\Delta$|lnL|<2) (Extended Data Fig. 7; Supplementary Table 7). In contrast to the C60 models, there is a major increase in phylogenetic signal in Poisson+CAT models in PhyloBayes towards Porifera-sister, and outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig. 7). For example, in Whelan2017_full matrix, we found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from 57.5% in matrices with distant outgroups (Holozoa) to 35.4% when outgroups are restricted (Choanimalia, Supplementary Table 7).

@@

#### Outgroup taxa sampling with C60 and CAT models
Because different choices of outgroups could also affect phylogenetic inference as suggested in previous analyses, we parsed the full data matrices into three different types of outgroups: Choanimalia, Holozoa and Opisthokonta. These datasets include the same set of genes but differ in the composition of outgroup species. Choanimalia only includes choanofagellates as outgroup; Holozoa also includes more distantly related Holozoans; Opistokonta also includes Fungi. For each Choanimalia data matrice, both C60 models in IQ-TREE and Poisson+CAT models in PhyloBayes were conducted. The maximum likelihood analysis was performed using the best-fit substitution model identified as above and nodal support was assessed with 1000 ultrafast bootstrap replicates using IQ-TREE. Moreover, Bayesian inference with the site-heterogeneous Poisson+CAT model was done with PhyloBayes-MPI v1.8. To minimize computational burden, GTR+CAT models were only performed in the representative Choanimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.

@@

#### Phylogenetic distribution of support
To investigate the distribution of phylogenetic signal of the animal-root position in data matrices, we considered three major data matrices from three studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013_est, and Whelan2017_full data matrices. We examined two hypotheses: Ctenophora-sister (T1) and Porifera-sister (T2) to the rest of metazoans, under a panel of evolutionary models with different outgroup schemes (Choanimalia and the full matrix). For IQ-TREE analysis in each data matrix, site-wise likelihood scores were inferred for both hypotheses using IQ-TREE (option -g) with the LG+G4 model. The two different phylogenetic trees passed to IQ-TREE 1.6.12 (via -z) were the tree where Ctenophora-sister and a tree modified to have Porifera placed as the sister to the rest of animals. The numbers of genes and sites supporting each hypothesis were calculated from IQ-TREE output and Perl scripts from a previous study [@shen2017contentious]. By calculating gene-wise log-likelihood scores between T1 and T2 for every gene, we considered a gene with an absolute value of log-likelihood difference of two as a gene with strong (|ΔlnL| > 2) or weak (|ΔlnL| < 2) phylogenetic signal as done in a previous study [@smith2020phylogenetic].

@@

#### Sensitivity analyses with different number of substitutional categories
To explore how the number of substitutional categories may affect the phylogenetic inference related to the animal phylogeny, we conducted PhyloBayes analyses with a panel of different substitutional categories in the Whelan2017_strict (ncat=60, 70, 80, 90, 110, 120, 150, 180, 360), Whelan2017_full_Choanimalia (nCAT=60, 340, 380, 420, 460), Philippe2009 (nCAT=60, 90, 120, 150, 180) Philippe2009_Holozoa (nCAT=360, 400, 440, 480) and Ryan2013_est (ncat=60). To compare the results between Poisson+CAT and GTR+CAT and minimize computational burden, GTR+CAT and GTR+CAT60 models were only performed in the representative Choanimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full. All PhyloBayes analyses were carried out using the same settings as above (see Outgroup taxa sampling with C60 and CAT models section), except when a different number of categories was used.

To compare the allocation of frequency categories across sites in the Philippe2009_Choanimalia and Whelan2017_strict matrices for the constrained Poisson+CAT60 model and unconstrained Posson+CAT model, we parsed the information of PhyloBayes chain files by sampling one in every 1000 generations after burnin determined above. The scripts and subsampled chain files are in the `../trees_new/frequency` subdirectory of the git repository.

#### Cross-validation analyses
Bayesian cross-validation implemented in PhyloBayes-MPI was used to compare the fit of Poisson+nCAT60 and Poisson+CAT models coupled with a gamma distribution of site-rate heterogeneity in Whelan2017_strict and Philippe2009_Choanimalia data matrices. Ten replicates were run, each replicate consisting of a random subsample of 10,000 sites for training the model and 2,000 sites for computing the cross-validation likelihood score. For each  run, 5,000 generations were run and the first 2,000 generations were discarded as burn-in.

@@

ggsave("figures/Extended_Data_Fig2_raw.pdf", width=10, height=8)
```
![](figures/Extended_Data_Fig2_manuscript.png)
**Extended Data Fig. 2.** Comparison of the tree topologies under different substitutional models in Philippe2009_Choanimalia data matrix. (A). WAG+C60. (B). Poisson+nCAT60. (C). Poisson+nCAT150. (D). Poisson+CAT. (E). GTR+CAT. All trees were visualized and annotated in R package ggtree [@yu2018two].

@@

**Extended Data Fig. 5.**  Box plots of log likelihood score for 10-fold cross-validation between Poisson+CAT and Poisson+nCAT60 models in Whelan2017_strict and Philippe2009_Choanimalia data matrices.

@@

# Generate plots
gg_rank_phil =
  pb_summaries %>%
  filter( matrix=="Philippe2009" ) %>%
  ggplot() +
  geom_line(aes(x=rank, y=count, col=model)) +
  scale_colour_manual(values = edf2_cols) +
  xlab( "category rank" ) +
  ylab( "site count" ) +
  theme(legend.position="right") + ggtitle("Philippe2009_Choanimalia")

@@

**Extended Data Fig. 6** The allocation of frequency categories across sites in the Philippe2009 matrix (left column) and Whelan2017_strict matrix (right column) for the constrained Posson+nCAT60 model and unconstrained Posson+CAT models (differentiated with color). These plots are for singe sample isolated from the end of PhyloBayes chain files. (A,B) The count of sites allocated to each equilibrium frequency category, with the categories ranked from the most abundant to least abundant along the x axis. The unconstrained CAT analyses have a long tail of categories that are allocated to very few sites, which adds a considerable number of parameters that pertain to only a small fraction of the data. The nCAT60 analyses, which are constrained to 60 sites, have no such long tail and the rarest categories are allocated to a far greater number of sites than any of the sites in the long tail of the CAT analyses.

@@

**Extended Data Fig. 7** The distribution of phylogenetic signal for Ctenophora-sister and Porifera-sister with different models and outgroup choice in Philippe2009, Ryan2013 and Whelan2017_full matrices (with two outgroup sampling: Choanimalia and full matrices). The two alternative topological hypotheses are: Ctenophora-sister; T1 (Orange); Porifera-sister T2 (Green). Proportions of genes supporting each of two alternative hypotheses in the Philippe2009, Ryan2013_est and Whelan2017_full data matrices with different outgroups sampling and substitutional models. The GLS values for each gene in each data matrix are provided in Table S5. We considered a gene with an absolute value of log-likelihood difference of two as a gene with strong (|ΔlnL| $\geq$ 2) or weak ( 0 < |ΔlnL| < 2) phylogenetic signal.

@@

Pick *et al.* [@Pick:2010eb] sought to test whether Ctenophora-sister was an artefact of insufficient taxon sampling. They added new and additional published sequence data to the 64-taxon matrix of Dunn *et al.* [@Dunn:2008ky]. The new taxa included 12 sponges, 1 ctenophore, 5 cnidarians, and *Trichoplax*. They further modified the matrix by removing 2,150 sites that were poorly sampled or aligned. They considered two different sets of outgroups: Choanoflagellata (resulting in Choanimalia) and the same sampling as Dunn *et al.* (resulting in Opisthokonta).

All their analyses were conducted under the F81+CAT+Gamma model in PhyloBayes, in both a Bayesian framework and with bootstrapping. All analyses have the same ingroup sampling and site removal so it isn't possible to independently assess the impact of these factors. Analyses with Choanimalia sampling recovered Porifera-sister with 72% posterior probability (PP) and 91% bootstrap support (BS). With broader Opisthokonta sampling, support for Porifera-sister is 84% PP. This is an interesting case where increased outgroup sampling leads to increased support for Porifera-sister.

@@

Whelan *et al.* [@whelan2017ctenophore] added 27 new ctenophore transcriptomic data to explore animal-root position as well as relationships within Ctenophores. It significantly increased ctenophore taxon sampling than other studies. Putative orthologs were determined largely similar to Whelan2015. The subsequent filtering strategy was also similar to the previous study. All analyses using site-homogeneous and site-heterogeneous models strongly supported Ctenophora-sister hypothesis, even with GTR+CAT model in Choanimalia dataset. The main conclusions of this study were based on Whelan2017_full and Whelan2017_strict matrices.

>>>>

![](figures/Figure_overview.png)
**Fig. 1.** Two alternative phylogenetic hypotheses for the root of the animal tree. (A) The Ctenophora-sister hypothesis posits that there is a clade (designated by the orange node) that includes all animals except Ctenophora, and that Ctenophora is sister to this clade. (B) The Porifera-sister hypothesis posits that there is a clade (designated by the green node) that includes all animals except Porifera, and that Porifera is sister to this clade. Testing these hypotheses requires evaluating the support for each of these alternative nodes. (C) The animals and their outgroup choice, showing the three progressively more inclusive clades Choanozoa, Holozoa, and Opisthokonta. (D) A ctenophore, commonly known as a comb jelliy. (E) A poriferan, commonly known as a sponge.

@@

Sampling of outgroup taxa (non-animals, in this case) is critical to phylogenetic rooting questions, since the node where the outgroup subtree attaches to the ingroup subtree is the root of the ingroup. There has therefore been extensive focus on improving outgroup sampling when testing phylogenetic hypotheses about rooting [@graham2002rooting]. Most studies addressing the animal root have removed more distantly related outgroup taxa in some analyses to explore the effect of outgroup selection to ingroup topology [@ryan2013genome; @pisani2015genomic]. Three progressively more inclusive clades have often been investigated: Choanozoa (animals plus most closely related Choanoflagellata), Holozoa (Choanozoa plus more distantly related Holozoa), and Opisthokonta (Holozoa + Fungi).

@@

Our summary of previous phylogenetic analyses (Fig. 3A) shows that Ctenophora-sister is supported in analyses that span the full range of outgroup sampling and models used to date, including some analyses with restricted outgroup sampling and models that accommodate site-heterogeneous equilibrium frequencies with CAT. This is consistent with previous assessments of the problem [@whelan2016let], but is drawn from a much more extensive and systematic examination. The only analyses that support Porifera-sister have reduced outgroup sampling (Choanozoa, Holozoa) and site-heterogeneous models with CAT. Model adequacy assessments generally favor GTR+CAT over Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], but because GTR+CAT is so parameter rich, many analyses that use a model with GTR+CAT do not converge. The fact that Porifera-sister is recovered only for particular models with particular outgroup sampling indicates that model and outgroup interact, and that this interaction is fundamental to understanding the range of results obtained across analyses.

@@

For each of the published studies, we selected the matrix that was the primary focus of the manuscript, or has been reanalyzed extensively in other studies, for further analysis. For each of these matrices, we progressively trimmed taxon sampling to create Opisthokonta, Holozoa, and Choanozoa versions, where permitted by original outgroup sampling. This produced `r analyses_new$matrix%>% unique()%>% length()` data matrices from 11 studies that presented new sequence data and for which partition data were available.

@@

We examined the specific impact of this large difference in category number on animal root position. It is currently not possible to use more than 60 categories for C models in IQ-TREE, but the number of categories can be set *a priori* in Phylobayes using nCAT. We therefore varied the number of categories in Phylobayes analyses (Fig. 4; Supplementary Table 5). We found that Poisson+nCAT60 analyses in PhyloBayes, like Poisson+C60 and WAG+C60 IQ-TREE analyses, provide strong support of Ctenophora-sister. This indicates that the difference in results between unconstrained CAT analyses in PhyloBayes and C60 analyses in IQ-TREE is not due to differences in the software or other implementation factors, but due to the large difference, in excess of ten-fold, in the number of site categories. When we increased the number of categories in PhyloBayes nCAT analyses, we observe the transition from support for Ctenophora-sister to an unresolved root to Porifera-sister (Fig. 4). For example, for the Whelan2017_strict matrix this transition occurs between 60 -- 120 categories when using Poisson+nCAT model. Due to computational limitations of GTR models, we only ran GTR+CAT and GTR+nCAT60 models on representative matrices with Choanozoa sampling. We found that for Whelan2017 matrices, support shifted from Porifera-sister with Poisson+CAT model to Ctenophora-sister using GTR+CAT model. Moreover, we also found all results strongly supported Ctenophora-sister with GTR+nCAT60 models.

These results further clarify when analyses support Porifera-sister (*e.g.* Whelan2017: Extended Data Fig. 1; Philippe2009: Extended Data Fig. 2): when outgroup sampling is restricted (Choanozoa), when a Poisson (rather than a GTR) exchange matrix is used, and when a very large number of site categories is used (unconstrained CAT, giving hundreds of equilibrium frequency categories). Analyses under other conditions either support Ctenophora-sister or are unresolved. This is consistent across published analyses and our new panels of analyses. The question, then, isn’t why similar analyses give different results, but how we should interpret variation in results when we run analyses that differ in these specific respects. If we fix the first two features to conditions that are necessary for Porifera-sister support (Choanozoa taxon sampling and a Poisson exchange matrix), there are several insights that we can glean from examining how the number of equilibrium frequency categories impacts results that sheds light on interpretation of those results.

The first questions regard the relative fit of adding aditional site-heterogeneous categories, and the adequacy of models with different numbers of categories. ModelFinder selects site heterogeneous C60 models (Supplementary Table 3), but IQ-TREE often gives a warning under C60 that the model may overfit, with too many categories, because some mixture weights are close to 0 (Supplementary Table 6). While ModelFinder considers the relative fit of models, this is distinct from model adequacy -- an absolute measure of how well the model can describe the variation in the data. Cross-Validation (CV) is an important tool to assess adequacy by examining the predictive power of the model, *i.e.* the ability of the model to predict observations [@lartillot2009phylobayes]. It could be that a model with a very large number of categories are required to adequately describe the site heterogeneity in these data, and that we should therefore favor the Porifera-sister results obtained in unconstrained CAT analyses. To evaluate the predictive performance and model-fit between Poisson+CAT and Poisson+nCAT60 models in Whelan2017_strict and Philippe2009_Choanozoa matrices, we conducted Bayesian cross-validation experiments with PhyloBayes. We found nearly identical distributions of cross-validation scores for Poisson+nCAT60 and Poisson+CAT models (Extended Data Fig. 5). A t-test analysis shows that there is no significant difference between them for either matrix (Whelan2017_strict: $p-value=`r P_value_Whelan2017_strict`$; Philippe2009_Choanozoa: $p-value=`r P_value_Philippe2009`$). This indicates that 60 categories in a Poisson+nCAT60 model are as adequate as at predicting the data as the hundreds of categories in an unconstrained Poisson+CAT model. This suggests there is no reason to favor the narrow analysis conditions under which Porifera-sister are obtained.

In a site-heterogeneous model, each site is allocated to one of the categories in each sample. Different categories can have very different numbers of sites. We examine category allocations in the last chain samples of analyses of the Philippe2009_Choanozoa and Whelan2017_strict matrices (Extended Data Fig. 6A-B), which have `r n_categories_phil_cat` and `r n_categories_whel_cat` categories, respectively. The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009_Choanozoa matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix. This long tail of categories that apply to a very small fraction of sites is in stark contrast to nCAT60 analyses, constrained to 60 categories, which have no such long tail of rare categories (Extended Data Fig. 6A-B). The fact that so many categories apply to such a small fraction of data be help explain why increasing the number of categories more than ten-fold has so little impact on the predictive power of these far more complex models.

@@

Although a higher proportion of genes support Ctenophora-sister with site-heterogeneous C60 models, the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models. In an extreme case, in matrices from Ryan2013_est nearly 30% of genes changed from strong $\Delta$|lnL|>2) to weak Ctenophora-sister signal ($\Delta$|lnL|<2) (Extended Data Fig. 7; Supplementary Table 7). In contrast to the C60 models, there is a major increase in phylogenetic signal in Poisson+CAT models in PhyloBayes towards Porifera-sister, and outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig. 7). For example, in Whelan2017_full matrix, we found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from 57.5% in matrices with distant outgroups (Holozoa) to 35.4% when outgroups are restricted (Choanozoa, Supplementary Table 7).

@@

#### Outgroup taxa sampling with C60 and CAT models
Because different choices of outgroups could also affect phylogenetic inference as suggested in previous analyses, we parsed the full data matrices into three different types of outgroups: Choanozoa, Holozoa and Opisthokonta. These datasets include the same set of genes but differ in the composition of outgroup species. Choanozoa only includes choanofagellates as outgroup; Holozoa also includes more distantly related Holozoans; Opistokonta also includes Fungi. For each Choanozoa data matrice, both C60 models in IQ-TREE and Poisson+CAT models in PhyloBayes were conducted. The maximum likelihood analysis was performed using the best-fit substitution model identified as above and nodal support was assessed with 1000 ultrafast bootstrap replicates using IQ-TREE. Moreover, Bayesian inference with the site-heterogeneous Poisson+CAT model was done with PhyloBayes-MPI v1.8. To minimize computational burden, GTR+CAT models were only performed in the representative Choanozoa matrices from Philippe2009, Ryan2013_est and Whelan2017_full.

@@

#### Phylogenetic distribution of support
To investigate the distribution of phylogenetic signal of the animal-root position in data matrices, we considered three major data matrices from three studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013_est, and Whelan2017_full data matrices. We examined two hypotheses: Ctenophora-sister (T1) and Porifera-sister (T2) to the rest of metazoans, under a panel of evolutionary models with different outgroup schemes (Choanozoa and the full matrix). For IQ-TREE analysis in each data matrix, site-wise likelihood scores were inferred for both hypotheses using IQ-TREE (option -g) with the LG+G4 model. The two different phylogenetic trees passed to IQ-TREE 1.6.12 (via -z) were the tree where Ctenophora-sister and a tree modified to have Porifera placed as the sister to the rest of animals. The numbers of genes and sites supporting each hypothesis were calculated from IQ-TREE output and Perl scripts from a previous study [@shen2017contentious]. By calculating gene-wise log-likelihood scores between T1 and T2 for every gene, we considered a gene with an absolute value of log-likelihood difference of two as a gene with strong (|ΔlnL| > 2) or weak (|ΔlnL| < 2) phylogenetic signal as done in a previous study [@smith2020phylogenetic].

@@

#### Sensitivity analyses with different number of substitutional categories
To explore how the number of substitutional categories may affect the phylogenetic inference related to the animal phylogeny, we conducted PhyloBayes analyses with a panel of different substitutional categories in the Whelan2017_strict (ncat=60, 70, 80, 90, 110, 120, 150, 180, 360), Whelan2017_full_Choanozoa (nCAT=60, 340, 380, 420, 460), Philippe2009 (nCAT=60, 90, 120, 150, 180) Philippe2009_Holozoa (nCAT=360, 400, 440, 480) and Ryan2013_est (ncat=60). To compare the results between Poisson+CAT and GTR+CAT and minimize computational burden, GTR+CAT and GTR+CAT60 models were only performed in the representative Choanozoa matrices from Philippe2009, Ryan2013_est and Whelan2017_full. All PhyloBayes analyses were carried out using the same settings as above (see Outgroup taxa sampling with C60 and CAT models section), except when a different number of categories was used.

To compare the allocation of frequency categories across sites in the Philippe2009_Choanozoa and Whelan2017_strict matrices for the constrained Poisson+CAT60 model and unconstrained Posson+CAT model, we parsed the information of PhyloBayes chain files by sampling one in every 1000 generations after burnin determined above. The scripts and subsampled chain files are in the `../trees_new/frequency` subdirectory of the git repository.

#### Cross-validation analyses
Bayesian cross-validation implemented in PhyloBayes-MPI was used to compare the fit of Poisson+nCAT60 and Poisson+CAT models coupled with a gamma distribution of site-rate heterogeneity in Whelan2017_strict and Philippe2009_Choanozoa data matrices. Ten replicates were run, each replicate consisting of a random subsample of 10,000 sites for training the model and 2,000 sites for computing the cross-validation likelihood score. For each  run, 5,000 generations were run and the first 2,000 generations were discarded as burn-in.

@@

ggsave("figures/Extended_Data_Fig2_raw.pdf", width=10, height=8)
```
![](figures/Extended_Data_Fig2_manuscript.png)
**Extended Data Fig. 2.** Comparison of the tree topologies under different substitutional models in Philippe2009_Choanozoa data matrix. (A). WAG+C60. (B). Poisson+nCAT60. (C). Poisson+nCAT150. (D). Poisson+CAT. (E). GTR+CAT. All trees were visualized and annotated in R package ggtree [@yu2018two].

@@

**Extended Data Fig. 5.**  Box plots of log likelihood score for 10-fold cross-validation between Poisson+CAT and Poisson+nCAT60 models in Whelan2017_strict and Philippe2009_Choanozoa data matrices.

@@

# Generate plots
gg_rank_phil =
  pb_summaries %>%
  filter( matrix=="Philippe2009" ) %>%
  ggplot() +
  geom_line(aes(x=rank, y=count, col=model)) +
  scale_colour_manual(values = edf2_cols) +
  xlab( "category rank" ) +
  ylab( "site count" ) +
  theme(legend.position="right") + ggtitle("Philippe2009_Choanozoa")

@@

**Extended Data Fig. 6** The allocation of frequency categories across sites in the Philippe2009_Choanozoa matrix (left column) and Whelan2017_strict matrix (right column) for the constrained Posson+nCAT60 model and unconstrained Posson+CAT models (differentiated with color). These plots are for singe sample isolated from the end of PhyloBayes chain files. (A,B) The count of sites allocated to each equilibrium frequency category, with the categories ranked from the most abundant to least abundant along the x axis. The unconstrained CAT analyses have a long tail of categories that are allocated to very few sites, which adds a considerable number of parameters that pertain to only a small fraction of the data. The nCAT60 analyses, which are constrained to 60 sites, have no such long tail and the rarest categories are allocated to a far greater number of sites than any of the sites in the long tail of the CAT analyses.

@@

**Extended Data Fig. 7** The distribution of phylogenetic signal for Ctenophora-sister and Porifera-sister with different models and outgroup choice in Philippe2009, Ryan2013 and Whelan2017_full matrices (with two outgroup sampling: Choanozoa and full matrices). The two alternative topological hypotheses are: Ctenophora-sister; T1 (Orange); Porifera-sister T2 (Green). Proportions of genes supporting each of two alternative hypotheses in the Philippe2009, Ryan2013_est and Whelan2017_full data matrices with different outgroups sampling and substitutional models. The GLS values for each gene in each data matrix are provided in Table S5. We considered a gene with an absolute value of log-likelihood difference of two as a gene with strong (|ΔlnL| $\geq$ 2) or weak ( 0 < |ΔlnL| < 2) phylogenetic signal.

@@

Pick *et al.* [@Pick:2010eb] sought to test whether Ctenophora-sister was an artefact of insufficient taxon sampling. They added new and additional published sequence data to the 64-taxon matrix of Dunn *et al.* [@Dunn:2008ky]. The new taxa included 12 sponges, 1 ctenophore, 5 cnidarians, and *Trichoplax*. They further modified the matrix by removing 2,150 sites that were poorly sampled or aligned. They considered two different sets of outgroups: Choanoflagellata (resulting in Choanozoa) and the same sampling as Dunn *et al.* (resulting in Opisthokonta).

All their analyses were conducted under the F81+CAT+Gamma model in PhyloBayes, in both a Bayesian framework and with bootstrapping. All analyses have the same ingroup sampling and site removal so it isn't possible to independently assess the impact of these factors. Analyses with Choanozoa sampling recovered Porifera-sister with 72% posterior probability (PP) and 91% bootstrap support (BS). With broader Opisthokonta sampling, support for Porifera-sister is 84% PP. This is an interesting case where increased outgroup sampling leads to increased support for Porifera-sister.

@@

Whelan *et al.* [@whelan2017ctenophore] added 27 new ctenophore transcriptomic data to explore animal-root position as well as relationships within Ctenophores. It significantly increased ctenophore taxon sampling than other studies. Putative orthologs were determined largely similar to Whelan2015. The subsequent filtering strategy was also similar to the previous study. All analyses using site-homogeneous and site-heterogeneous models strongly supported Ctenophora-sister hypothesis, even with GTR+CAT model in Choanozoa dataset. The main conclusions of this study were based on Whelan2017_full and Whelan2017_strict matrices.
